---
title: "Simulation_bio"
author: "Joonwon Lee"
date: "2023-12-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(tidyverse)
```

# Update 12/4

```{r}
setwd("C:/Users/joonw/Downloads")
data <- read.csv("pilot.csv")
```


```{r}
df <- data %>%
  group_by(subject) %>%
  mutate(diff_y = y[1]-y)

df = df[df$diff_y!=0,]
df = df %>%
  group_by(subject) %>%
  mutate(subject_mean = mean(diff_y), tmp = (diff_y-subject_mean)^2)

total_mean_y = mean(df$diff_y)
sub_mean_y = unique(df$subject_mean) 
between_var_y = sum((sub_mean_y- total_mean_y)^2)/ (length(sub_mean_y)-1)
within_var_y =  sum(df$tmp)/dim(df)[1]  

mean_diff_y = sub_mean_y  
between_sd_y = sqrt(between_var_y)         
within_sd_y = sqrt(within_var_y)      


df2 <- data %>%
  group_by(subject) %>%
  mutate(diff_z = z[1]-z)

df2 = df2[df2$diff_z!=0,]
df2 = df2 %>%
  group_by(subject) %>%
  mutate(subject_mean = mean(diff_z), tmp = (diff_z-subject_mean)^2)

total_mean_z = mean(df2$diff_z)
sub_mean_z = unique(df2$subject_mean) 
between_var_z = sum((sub_mean_z- total_mean_z)^2)/ (length(sub_mean_z)-1)
within_var_z = sum(df2$tmp)/dim(df2)[1]

mean_diff_z = sub_mean_z  
between_sd_z = sqrt(between_var_z)        
within_sd_z = sqrt(within_var_z)  
```

## Summary table of y and z

```{r}
## y
(Grand_mean = mean(df$diff_y))
(subject_1_mean = sub_mean_y[1])
(subject_2_mean = sub_mean_y[2])
(subject_3_mean = sub_mean_y[3])
(subject_4_mean = sub_mean_y[4])
(between_subject_variance = var(sub_mean_y))
(Within_subject_variance = within_var_y)

## z

(Grand_mean = mean(df2$diff_z))
(subject_1_mean = sub_mean_z[1])
(subject_2_mean = sub_mean_z[2])
(subject_3_mean = sub_mean_z[3])
(subject_4_mean = sub_mean_z[4])
(between_subject_variance = var(sub_mean_z))
(Within_subject_variance = within_var_z)

```





# Questions 1: how many data points needed in each group?
## One sample T test

Remark: 
Basically, for each subject, we have (j+1) data points:
\[
ov-cycle_1, \ ov-cycle_2, \  \cdots, \ ov-cycle_j. 
\]
But ov is fixed, which means it has no randomness within the subject. Randomness comes from other j stages in the cycle except ovulation stage. So we can design the problem as one sample T test and our hypothesis can be written as: 
$H_0$: mean(ov_cycle_i) = 0 VS  $H_a$: mean(ov_cycle_i)=1,

where $1 \leq i \leq j$.

```{r}
mean_diff = c(0.3,0.4,0.5)
within_sd = c(0.2,0.25,0.3)
sigma = within_sd[2]
diff = mean_diff[1]
power.t.test(sig.level = 0.05, power= 0.9, delta= diff, sd= sigma, type="one", alternative="one.sided")
```

# Question2: How many subjects do we need assuming each subject has 8 stages?
## Power analysis by simulation
Consider following model and a hypothesis test:
\[
y_{ij} = \mu + b_i + e_{ij}, \ b_i \sim N(0,\sigma_1^2), \ e_{ij} \sim N(0, \sigma_2^2),
\]

One of the reasons why we are not using "session + estrogen+ progesterone" is that they seem to be
correlated. Even if they are not linearly correlated, their correlation may cause the variance to increase. 

$H_0: \mu = 0$, VS $H_a: \mu \neq 0 $.


### Model1, check if 10 subjects are enough.

Power of a test: probability that the test will correctly reject a false null hypothesis.
When estimating $\mu$, I think degrees of freedom for this is (n-1) as df for residual but not so sure. Then I put df = n- k(subject numbers) in model 2.

```{r}
pow.sim = function(reps=1000, seed=2023, mu = 0.3, n=10, ov_cycle=7, between_sd = 0.15, within_sd = 0.15){
  set.seed(seed)
  sig = 0
  capture = rep(0,reps)
  
  for (i in 1:reps){
  bi = rnorm(n, mean=0, sd = between_sd)
  b = rep(bi, each = ov_cycle)
  e = rnorm(ov_cycle*n, mean=0, sd=within_sd)
  y = mu + b + e

  subjects = factor( rep(1:n, each = ov_cycle))
  data = data.frame(y=y, group = subjects )
  mod1 = lmer(y~ 1+ (1|group), data=data)

  # set significance level alpha = 0.05 and do two sided hypothesis test
  # H_0 : mu = 0,  H_a: mu \neq 0
  critical_val = qt(1-0.05/2, df= (n*ov_cycle-1 ))[1]
  # Extract coefficients and standard errors
  coefficients <- fixef(mod1)
  se <- sqrt(diag(vcov(mod1)))
  # Calculate t-values
  t_values <- coefficients / se
  
  sig = sig + ifelse( t_values >critical_val, 1,0)
  }
  return (sig/reps)
}
```

```{r}
within_sd = c(0.2, 0.25, 0.3)
between_sd = c(0.2,0.25,0.3)
mu = c(0.3,0.4,0.5)
sub_num = c(10,12)

# n=10
test = matrix(0,3,3)
for (i in 1:3){
  for (j in 1:3){
    test[i,j] = pow.sim(reps=1000, seed=2023, mu = mu[i], n=sub_num[1], ov_cycle=7, between_sd = within_sd[j], within_sd = between_sd[j])
  }
}

# n= 12
test2 = matrix(0,3,3)
for (i in 1:3){
  for (j in 1:3){
    test2[i,j] = pow.sim(reps=1000, seed=2023, mu = mu[i], n=sub_num[2], ov_cycle=7, between_sd = within_sd[j], within_sd = between_sd[j])
  }
}
```

It seems 10 subjects are sufficient.

### Update pilot 4

```{r}
print( sqrt(c(within_var_y,within_var_z)))
within_sd_y = c(0.14, 0.2, 0.25, 0.3)

within_sd = within_sd_y
between_sd = within_sd_y
mu = sort(abs(mean_diff_y), decreasing=FALSE)  #  0.08, 0.27, 0.53, 0.76

sub_num = c(10,12)

# n=10
test_y = matrix(0,4,4)
for (i in 1:4){
  for (j in 1:4){
    test_y[i,j] = pow.sim(reps=1000, seed=2023, mu = mu[i], n=sub_num[1], ov_cycle=7, between_sd = within_sd[j], within_sd = between_sd[j])
  }
}
within_sd_z = c(0.23, 0.3, 0.35, 0.4)

within_sd = within_sd_z
between_sd = within_sd_z
mu = sort(abs(mean_diff_z), decreasing=FALSE)  #  0.14, 0.45, 0.76, 1.116
  
test_z = matrix(0,4,4)
for (i in 1:4){
  for (j in 1:4){
    test_z[i,j] = pow.sim(reps=1000, seed=2023, mu = mu[i], n=sub_num[1], ov_cycle=7, between_sd = within_sd[j], within_sd = between_sd[j])
  }
}

```

### Model2, check if 10 subjects are enough. Actually model2 and model1 is equivalent. 

Consider following model2:
\[
y_ij = \mu + \beta O_{ij} + b_i + e_{ij},
\]
where $O_{ij}$ is an indicator variable taking value 1 if ovulation in a subject and 0 otherwise.

Here we are testing $H_0: \beta = 0.$

 


```{r}
pow.sim2 = function(reps=1000, seed=2023, mu = 0.3, n, ov_cycle=8, between_sd = 0.15, within_sd = 0.15, beta=2){
  set.seed(seed)
  sig = 0
 

  for (i in 1:reps){
  b_i = rnorm(n, sd = between_sd)
  b = rep(b_i, each = ov_cycle)
  tmp = c(1, rep(0,7))
  o = rep(tmp, n) 
  e = rnorm(ov_cycle*n, 0, sd=within_sd)
  y = mu + beta*o + b + e
  subjects = factor( rep(1:n, each = ov_cycle))
  data = data.frame(y=y, group = subjects )
  mod2 = lmer(y~ 1+ o + (1|group), data=data)  
    
  # set significance level alpha = 0.05 and do two sided hypothesis test
  # H_0 : mu = 0,  H_a: mu \neq 0
  critical_val = qt(1-0.05/2, df= (n*ov_cycle-1 ))[1]
  # Extract coefficients and standard errors
  coefficients <- fixef(mod1)
  se <- sqrt(diag(vcov(mod1)))
  # Calculate t-values
  t_values <- coefficients / se
  
  sig = sig + ifelse( t_values >critical_val, 1,0)
  }
  return (sig/reps)
}
```

```{r}
within_sd = c(0.2, 0.25, 0.3)
between_sd = c(0.2,0.25,0.3)
mu = c(0.3,0.4,0.5)
sub_num = c(10,12)
beta = 2

# n=10
test = matrix(0,3,3)
for (i in 1:3){
  for (j in 1:3){
    test[i,j] = pow.sim2(reps=1000, seed=2023, mu = mu[i], n=10, ov_cycle=8, between_sd = between_sd[j], within_sd = within_sd[j],beta=2)
  }
}

# n= 12
test2 = matrix(0,3,3)
for (i in 1:3){
  for (j in 1:3){
    test2[i,j] = pow.sim2(reps=1000, seed=2023, mu = mu[i], n=12, ov_cycle=8, between_sd = within_sd[j], within_sd = between_sd[j], beta=2)
  }
}
```
